Differential effects of knapping skill acquisition on the cultural reproduction of Late Acheulean handaxe morphology: Archaeological and experimental insights

### 2023-02-05 manuscript/manuscript_word.Rmd
substantial revision after reframing.

Although we are not the first research team to use secondary archaeological data [e.g., @keyHandaxeShapeVariation2019], we would still like to highlight here that this research project further exemplifys the potential of reusing old archaeological data in digital format to address novel research questions. In this paper, the main source of archaeological data is a collection of photos produced and curated more than 10 years ago, and the morphological variation data of the experimental collection are also derived from photographs instead of remeasurements of the original artifacts. Given the irreversible nature of archaeological excavations, digitized data, be it text, pictures, or videos, often become the sole evidence that is available for certain research questions. Yet, it has been widely acknowledged that the reuse of archaeological data has not received enough attention among researchers in our discipline [@faniel2018; @huggett2018; @moody2021]. Among many reasons preventing archaeologists from reusing published and digitized data [@sobotkova2018], the lack of a standardized practice of and motivation for data sharing is a prominent one [@marwick2018]. As stated in the method section, we addressed this issue by sharing the raw data and the code for generating the derived data on an open-access repository. Another major and legitimate concern of archaeological data reuse is their quality. In terms of this aspect, we do acknowledge the limitations of relying on photos when it comes to the more detailed technological analysis of stone artifacts, however, our paper shows that finding the appropriate research questions given the data available is key to revealing new novel insights into the studied topic. Moreover, we believe that this type of research has a strong contemporary relevance due to the continued influence of the COVID-19 on fieldwork-related travel and direct access to archaeological artifacts [@balandier2022; @ogundiran2021].

@@

Regarding the research question we proposed in the beginning, our case study suggested that skill acquisition has a differential effect in the cultural reproduction of different aspects of mental templates, where cross-sectional thinning (PC1) is more constrained by knapping skill while elongation and pointedness (PC2) is less so. At a larger theoretical level it questions the distinction between social learning of design targets vs. individual learning of the skills needed to achieve them. Traditionally archaeological experiments speaking to the literature of cultural evolution tend to use handaxe as a model artifact and focus on how copying errors emerge during the transmission of a fixed and static target using transmission chain design and alternative raw materials such as foam [@schillinger2017; @schillingerConsideringRoleTime2014; @schillinger2015]. This line of inquiry is generally characterized by high internal validity (causal mechanisms) but low external validity (generalizability to archaeological data). In contrast, our study unpacks the differential reproductions of two major sources of variation and reveals how the development of motor skill during learning is constraining the achievement of the socially learnt design target, through an actualistic experimental setting featuring a higher degree of external validity [@liu2022]. In the future, more robust experimental studies are needed to deepen our understanding of the relationship between skill acquisition and the morphological variability of handaxes in the proper developmental context [@högberg2018; @nowell2021; @lew-levy2020] as well as their implications for the biological and cultural evolution of the hominin lineages.

>>>>

The negative correlation between the PC1 and PC2 values revealed a hidden structural constraint regarding the relationship between cross-sectional thinning and the imposed form. Our results (**Figure** \@ref(fig:PCcorrelation)) suggested thinner handaxes (low PC1 value) are generally more pointed/less ovate (high PC2 value), which was first reported in Crompton and Gowlett's [-@crompton1993] pioneering study on the allometry of Kilombe handaxes. In the thinning phase of handaxe making, a knapper must strike flakes that travel more than one half way across the surface while not breaking the handaxe into half [-@callahan1979: 90]. As a corollary, we speculate that it would be easier to perform thinning if the plan shape of a handaxe is narrower and more pointed, echoing the high technological difficulty of making large yet thin bifacial points as perceived by American hobbyist flintknappers [@whittaker2004: 180-182]. It is possible that such constraints help to explain why our novice knappers on average produced more handaxes in similar shapes to those preferred by modern expert knappers, however, this clearly does not explain the design target at Boxgrove. Given the ovate forms of the Boxgrove assemblage, it thus requires a high skill level to overcome this structural constraint to produce thin yet wide handaxes as demonstrated by the Boxgrove knappers. This also provides an alternative explanation to the cultural reproduction of form for the experimental convergence on pointed forms. In this comparative context, it would only be the Boxgrove assemblage that provided evidence of social conformity on a more difficult target shape.

As previously shown in Key's [-@keyHandaxeShapeVariation2019] previous finding regarding Boxgrove, it is also noteworthy how constrained the range of Boxgrove assemblage morphological variation is as measured by both PC1 and PC2 even when compared with the modern expert group (**Figure** \@ref(fig:GeneralPCA1)), especially given the fact that it has the largest sample size among all studied groups. Some potential explanations for this phenomenon include 1) the strong idiosyncrasy of individual expert knappers shaped by their own unique learning and practice experience; 2) the present-day skill shortage of our expert knapper as compared with Boxgrove knappers despite their multiple years of knapping practice [@milks2019]; and/or 3) modern knappers' skill level was affected by time constraints when they were requested to produce the reference collections [@schillingerConsideringRoleTime2014; @lewis2022].

@@

Regarding the research question we proposed in the beginning, our case study suggested that the processes of skill learning in a lithic medium do exert biases on the cultural reproduction of artifact morphology. More specifically, skill acquisition has a differential effect on the cultural reproduction of different aspects of mental templates, where cross-sectional thinning (PC1) is more constrained by knapping skill while elongation and pointedness (PC2) is less so. At a larger theoretical level, it questions the distinction between social learning of design targets vs. individual learning of the skills needed to achieve them. In the future, more robust experimental studies are needed to deepen our understanding of the relationship between skill acquisition and the morphological variability of handaxes in the proper developmental context [@högberg2018; @nowell2021; @lew-levy2020] as well as their implications for the biological and cultural evolution of the hominin lineages.
